---
title: "Week3Assignment"
author: "STUART.T"
date: "2024-08-01"
output:
  pdf_document: default
---

```{r setup, include=FALSE, echo= TRUE, message = FALSE}
#setup code for RMD, from the basic Rmd template provided in Rmd new documents. 
#Not my code.
knitr::opts_chunk$set(echo = FALSE)
```


my code:
```{r head, echo= TRUE, message= FALSE}

#Week 3 Assignment - Data Science as a Field


#imports
library(tidyverse)
library(lubridate)
library(ggplot2)
#library(scales)
library(dplyr)

#NYPD Shooting Incident Data
shooting.data.url <- 
  "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

#read to a dataframe
shooting.data <- read_csv(shooting.data.url)

#show summary
#note that some columns have data that is not useful for visualization. 
#for example, the case ID or "incident_key" number does not need to be analyzed
#precint numbers might be useful, but as categorical data
#summary(shooting.data)
#taking a look at the first few columns
#head(shooting.data)


```


## Main Code Block
### Includes 2 visualizations and a least-squares fitted line over a bar plot

```{r main, echo = TRUE}

#visualizations
#suspect by age group
#I notice several apparent mistakes in the data, since "1028" is not a valid age
age.data.clean <- shooting.data[!is.na(shooting.data$PERP_AGE_GROUP), ]
#removing any manually added "null" values 
age.data.clean <- age.data.clean[(age.data.clean$PERP_AGE_GROUP != "(null)"),]
#removing suspcious ages - found ages 940, 1020, 1028, and 224?
age.data.clean <- 
  age.data.clean[!(age.data.clean$PERP_AGE_GROUP %in% c(940, 1020, 1028, 224)),]
suspect.age <- table(age.data.clean$PERP_AGE_GROUP)
#a pie chart breakdown of ages
pie(suspect.age)
#suspect.age



#victim by race
age.data.clean <- shooting.data[!is.na(shooting.data$VIC_RACE), ]

#removing any unknown values that might skew data 
age.data.clean <- age.data.clean[(age.data.clean$VIC_RACE != "UNKNOWN"),]
victim.race <- table(age.data.clean$VIC_RACE)

#bar plot showing the differences in victims
#this does not work for the labels
#barplot(victim.race, horiz = FALSE)
#trying again with smaller font:
barplot(victim.race, horiz = FALSE, cex.names = .35)
#victim.race


#model portion
## of shootings over time
#for data tidying, we only need to remove null dates. The fact that there was a 
#shooting is shown by the existence of a row
#shooting.data.sorted <- arrange(shooting.data, shooting.data$OCCUR_DATE)
shooting.data$OCCUR_DATE <- as.Date(shooting.data$OCCUR_DATE, 
                                    format = "%m/%d/%Y")
shooting.data.sorted <- shooting.data[order(shooting.data$OCCUR_DATE), ]

#pivoting to summarize by month. adding a column for "month" by 
#formatting the date object
shooting.data.sorted$month <- format(shooting.data.sorted$OCCUR_DATE, "%Y-%m")
monthly.data <- table(shooting.data.sorted$month)

#monthly.data.sorted <- order(monthly.data)
#a barplot as a sanity check. 
#note there appears to be seasonality to the dates of shootings...
#barplot(monthly.data)

#monthly.data
#this is what I want to model. 
#a more complex model would use time series analysis due to the clear 
#seasonality, but I will start with a basic linear model over time. 

#linear model using lm()
#monthly.data
#in order to regress, I need to turn the months into sequential numbers 1, 2,..n
X <- 1:length(monthly.data)

linear.model <- lm(monthly.data ~ X, data = monthly.data)
#summarizing the model
summary(linear.model)

plot(monthly.data, col = "black", main = 
       "Shootings by Month, with Fitted Least Squares Line")
abline(linear.model, col = "blue")

```

## Bias discussion
In any dataset containing age, race, and locality demographics, there is almost certainly bias in both the sampling of the data and in the reporting of it. However, one of the most obvious biases shown by the data is that there are more Black victims than all other races combined. This does not seem to come from sampling, but is an actual representation of the data. In other words, this is a example of unfortunate disparity in the real world, not a data reporting failure. 
There is also bias in data reporting, however, as many of the raw data columns are incomplete or have null values. While this may not be intentional, it does show that across many variables the data could be skewed in any direction. I have made sure to minimize any bias of my own to the greatest possible extent, by modeling shootings as a function of time without any other variables. However, if one were to predict using multiple variables such as time and race, or time and location, that might inadvertently imply causality where there is none. 

## Model Discussion
Worth exploring further is how seasonal the data seems to be. There was a noticeable spike in activity shortly after the Covid-19 lockdowns, but the seasonal element still held. There might be other factors not included in this dataset which could better explain this. Perhaps air temperature is a contributing factor, but that is outside the scope of this discussion. The overall trend appears to be downward, even when including the increased number of shootings after 2020. 

### Plagiarism Statement
All analysis, code, and ideas were my own. I used google search and the documentation for assistance with syntax as needed, but all code contained in this document is produced by the author. 

### required session information for reproducibility
```{r info, echo= TRUE}
sessionInfo()
```


